A Bayesian approach to uncertainty in word embedding bias estimation

### 2023-09-07 paper/CLsubmission/bayesBiasCL.tex
reviewed works related

\item \citet{Ethayarajh2020Bernstein} correctly argues that a bias estimate should not be expressed
  as a single number without taking into account that the estimate is
  made using a sample of data and therefore has intrinsic uncertainty.
  The authors suggest using Bernstein bounds to gauge the uncertainty in
  terms of confidence intervals. We do not discuss this approach
  extensively, as we think that confidence intervals are quite
  problematic for several reasons, among others the confusing
  interpretability. We do not think that Bernstein bounds provide the
  best solution to the problem. Applying this method to a popular
  WinoBias dataset leads to the conclusion that more than 11903 samples
  are needed to claim a 95\% confidence interval for a bias estimate.
  This amount vastly exceeds the existing word lists for bias
  estimation. We propose a more realistic Bayesian method. Our
  conclusion is still that the word lists are sometimes too small, but
  at least they allow for gauging uncertainty as we go on to improve our
  methodology and extend the lists gradually.

@@

\item \citet{Guo2021CEAT} introduce the Contextualized Embedding Association Test
  Intersectional, meant to apply to dynamic word embeddings and,
  importantly develop methods for intersectional bias detection. The
  measure is a generalization of the WEAT method. The authors do inspect
  a distribution of effect sizes that arises from the consideration of
  various possible contexts, but they continue to standardize the
  difference in averaged means and use a single-number summary: the
  weighted mean of the effect sizes thus understood. The method,
  admittedly, deserves further evaluation which goes beyond the scope of
  this paper.
\end{itemize}

>>>>

@@

\item \citet{Guo2021CEAT} introduce the Contextualized Embedding Association Test
  Intersectional, meant to apply to dynamic word embeddings and,
  importantly develop methods for intersectional bias detection. The
  measure is a generalization of the WEAT method. The authors do inspect
  a distribution of effect sizes that arises from the consideration of
  various possible contexts, but they continue to standardize the
  difference in averaged means and use a single-number summary: the
  weighted mean of the effect sizes thus understood. The method,
  admittedly, deserves further evaluation which goes beyond the scope of
  this paper.

\item \citet{Lum2022Debiasing} observe that many group-wise performance meta-metrics used in algorithmic fairness consideration are biased estimators of disparities and propose a double-corrected variance estimator, which provides unbiased estimates and uncertainty quantification of the variance of model performance. This is certaintly valuable. Our approach differs in the following dimensions: it is not clear how bias variance estimator for model performance should be used in the context of group-wise cosine similarity evaluation, where we are not dealing with performance, but with cosine similarities, and where we are not dealing with a binomial distribution. Moreover, while having an unbiased estimate is valuable, if we have one this only means that \emph{in the long run} our estimates would tend to the true value. In the current situation we were looking at, the datasets are relatively small and we want to focus on what can be said before the long run has passed. Moreover, our Bayesian approach allows for regularization, inspection at values levels of granularity, and more straightforward interpretability, as it results in posterior distributions.

\item \citet{Ethayarajh2019understanding} point out that WEAT will be overblown in degenerate cases: if the word groups are singletons, the effect size is always maximal in one direction. They also bring up  sensitivity to the word frequency in a given corpus and to the word list choice. The authors propose a measure they call Relational Inner Product Association (RIPA) to measure the association between a word vector and a relation vector in  a  word embedding. Roughly, this is  the scalar projection of a word vector onto a one-dimensional bias subspace (which is found by employing approach from \citet{Bolukbasi2016man}). They do not attempt any statistical analysis or generalization to a bias measure that would apply to an embedding space as a whole (as opposed to assigning RIPA to one word vector with respect to a relation vector), or a particular protected word list. In contrast, our analysis allows for a more general evaluation, with uncertainty estimates. Moreover, RIPA can be used only for those embedding models that implicitly employ matrix factorization and contain co-occurrence statistic. In contrast, such technicalities do not prevent the Bayesian approach from being deployed for contextual models.

\item \citet{Ethayarajh2020Bernstein} argues that a bias estimate should not be expressed as a single number without taking into account that the estimate is
made using a sample of data and therefore has intrinsic uncertainty.
The author suggest using Bernstein bounds to gauge the uncertainty in
terms of confidence intervals. We do not discuss this approach
extensively, as we think that confidence intervals are quite
problematic for several reasons, among others the confusing
interpretability. We do not think that Bernstein bounds provide the
best solution to the problem. Applying this method to a popular
WinoBias dataset leads to the conclusion that more than 11903 data points of protected words 
are needed to claim a 95\% confidence interval for a bias estimate.
This amount vastly exceeds the existing word lists for bias
estimation. We propose a more realistic Bayesian method. Our
conclusion is still that the word lists are sometimes too small, but
at least they allow for gauging uncertainty as we go on to improve our
methodology and extend the lists gradually.

\item \citet{zhang2020robustness} shows the limitations of methods that use gender word pairs to detect bias in an embedding space. They claim that using analogies to detect bias may not necessarily reflect societal bias, but rather simply co-occurrence frequency between words. They conduct experiments where they evaluate four popular bias mesures: Direct Bias (DB), Word Association (WA), Neighbourhood Bias Metric (NBM), and Relational Inner Product Association (RIPA). They show that these measures are not robust to changing either the base pair or the form of a word used. This is  a valid point, to some extent related to the limitations on non-contextual models, and to some extent suggesting the inclusion of various word forms in the word list, and not constructing base direction using small word lists. As one of our observations is that the uncertainty resulting from the use of currently existing protected word lists is too large to justify sweeping statements, this is in line with our criticism.

\item \citet{Du2021Assessing} investigate the reliability of bias measures. They find that key existing candidates for bias measure often fail to agree with one another on particular pairs and protected words, and are sensitive to the word embedding algorithm and the corpus used. While some of these sensitivities are not necessarily signs of failure, we agree that the more abstract a measure is the more degrees of freedom for particular ways it is constructed, and the more ways such measures can disagree without a clear and principled reason. This is partially why we propose a less abstract approach that estimates expected cosine distances directly using raw data points.

\item \citet{Goldfarb2021BiasNotCorrelate} further our  understanding of  the relationship between intrinsic bias measured as a property of an embedding space, and extrinsic bias, measured in terms of  downstream task performance. They compare  WEAT as an intrinsic measure with Equality of Opportunity and Predictive Parity as external metrics.  The conclusion is such a  correlation is very limited. We consider this to be a point that matches our criticism of WEAT. Running a similar analysis for the Bayesian approach we discussed in this paper would be a useful task, which remains beyond the scope of this paper. One interesting general difference is that our method often claims that there are no sufficient  reasons to claim that an embedding is biased, so one would be on the lookout for such cases in which extrinsic bias measures nevertheless suggest the presence of bias. Such a presence, however, would not necessarily have to mean that the Bayesian approach to estimating potential systematic differences in cosine similarity is wrong, but rather suggest that external bias in downstream performance is not a function thereof.

\item \citet{Spliethover2021BiasSilhouette} propose Bias Silhouette Analysis (BSA), a method for assessing the quality of metrics that measure bias in word embedding models based on word lists. The core idea here is to quantify how much the bias values of a metric vary depending on what words from the lists are actually observed, where the computations result in  values for each model obtained  using word list subsets of increasing length. This allows for an inspection of bias metric convergence and sensitivity to word list choice, with a biased (GloVe) and an explicitly debiased model whose lower bias has been confirmed empirically (NBatch).  They examine the Embedding Coherence Test (ECT), the Relative Negative Sentiment Bias (RNSB), and the Word Embedding Association Test (WEAT), concluding  that none of these metrics can reliably discriminate between biased and non-biased models in all cases. This is not unexpected. An interesting question is what would happen if a similar test was applied to our method. However, our point is that the existing word lists are too short to provide reliable estimate of bias.

\end{itemize}
